APOGEE CATALOG OF 10,000+ M DWARFS MEASURED BY DATA-DRIVEN MODELS

### 2018-07-19 paper/draft1/draft1.tex
reorganize files
draft backup

\title{DATA DRIVEN MODELS FOR APOGEE M DWARFS}

@@

The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively. Finally we compare our temperature/metallicty model to theoretical models of M dwarf atmospheres (PHOENIX, \citealt{Husser:2013}; BTSETTL, \citealt{Allard:2011}) and identify features of missing line strengths.

@@

% \begin{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.

%define 'labels' - could be physical parameters or more empirical

\begin{enumerate}

\item[-] Define what parameters/classifications are important for Mdwarfs. We want to estimate physical parameters (such as Teff, logg, metallicity) and empirical parameters/classifications (SPT, etc.), which we will refer to as \emph{stellar labels}.

\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources... Terrien, Mann, Newton, Dressing catalogs)

\item[-] brief importance statment about APOGEE - new instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available

\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity). \citealt{Gilhool:2018}; \citealt{Skinner:2018}.

\item[-] Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.

@@

\subsection{Training Samples}

The Cannon is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}). The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not. Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier to infer from.

@@

% \begin{figure}[ht]
% \plottwo{figures/west_train_dist.png}{figures/mann_train_dist.png}
% \caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest} \label{fig:train_dist}
% \end{figure}

@@

APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample .

Describe ASPCAP spectra files and Ness continuum normalization procedure.

@@

\begin{enumerate}
\item[-] Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
\item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
\item[-] Briefly describe model inputs and outputs (1D spectral type model and 2D teff/metallicity and label vectors for those). Mention other models we've tested.
\end{enumerate}

Constructing such a model requires two steps: first the \emph{training step} in which the generative model is constructed at each pixel from the set of reference labels and fluxes, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum

% The key advantages of The Cannon being:
% Data-driven models take away the challenge of directly infering labels from a survey, instead we \emph{transfer labels} from another survery that is more accurate or easier to model.

Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a model that assumes the flux at each pixel of the spectrum is parameterized as a function \textit{g} 
\begin{equation}
	f_{n\lambda} = g(\ell_n | \theta_{\lambda}) + {\rm noise}
\end{equation}

\begin{equation}
	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + [s_{\lambda}^{2} + \sigma_{n\lambda}^{2}] \,\xi_{n\lambda}
\end{equation}

Quadratic label vectors:
\begin{equation}
	\ell_{n} = [1, SPT, SPT^{2}]
\end{equation}

\begin{equation}
	\ell_{n} = [1, T_{\rm eff}, [Fe/H], T_{\rm eff} \cdot [Fe/H], [Fe/H]^{2}]
\end{equation}

@@

\subsection{Spectral Type Model}

\begin{enumerate}
\item[-] We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure  displays 
\end{enumerate}

\begin{enumerate}
\item[-] For the physical parameter model we trained The Cannon on 30 M dwarfs with a two-dimensional with temperature/metallicity labels, and was able to achieve precisions of 44K/0.05dex respectively, more precise than both of the original training label uncertainties of 60K/0.08dex.

\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays

\item[-] Vsini and logg were not included as training parameters. Vsini is less import of a parameter as the sample consists of mainly old, slowly rotating sources (check what are ages?). Logg because?
\end{enumerate}

\begin{enumerate}
\item[-] Describe derivative plots and jacknife error estimation.
\end{enumerate}

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=18cm]{figures/self_and_validation_test_spt.png}
\end{center}
\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=18cm]{figures/self_and_validation_test_teff.png}
\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h.png}
\end{center}
\caption{Temperature and Metallicity label self test.} \label{fig:mann_selftest}
\end{figure}

\begin{figure}[ht]
\plotone{figures/1_Spectral_Sequence.pdf}
\plotone{figures/2_Spectral_Sequence.pdf}
\plotone{figures/3_Spectral_Sequence.pdf}
\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
\end{center}
\caption{\textit{Top two panels:} Mann-trained model for varying temperatures.} \label{fig:demo_teff}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_feh1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh3.png}
\end{center}
\caption{\textit{Top two panels:} Mann-trained model for varying metallicities.} \label{fig:demo_feh}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/derivative_jackknife_spt.png}
\end{center}
\caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$.} \label{fig:west_derivative}
\end{figure}

@@

@@

% \begin{longrotatetable}
\startlongtable
\begin{deluxetable*}{ccc|ccc|ccc}
\tablecaption{Cannon results for Temperature/Metallicity model \label{mann_results}}
\tabletypesize{\scriptsize}
\tablehead{
% \colhead{Designation} & \colhead{RA} & \colhead{DEC} & \colhead{$T_{\rm eff}$} & \colhead{[Fe/H]} \\
% \colhead{(2MASS)} & \colhead{(deg)} & \colhead{(deg)} & \colhead{(K)} & \colhead{(dex)}
\multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}  
} 
\startdata
2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3532       & 3519        & -0.3       & -0.27     & -0.24      \\
2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3527       & 3554        & -0.3       & -0.28     & -0.27      \\
2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3185       & 3188        & 0.11       & 0.05      & 0.02       \\
2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3783       & 3778        & 0.04       & 0.11      & 0.13       \\
2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3094       & 3101        & -0.07      & -0.01     & 0.03       \\
2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3218       & 3217        & 0.1        & 0.03      & 0.0        \\
2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3057       & 3074        & -0.16      & -0.1      & -0.01      \\
2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3249       & 3253        & -0.12      & -0.29     & -0.29      \\
2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3468       & 3467        & -0.12      & -0.26     & -0.27      \\
2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3381       & 3369        & -0.35      & -0.25     & -0.18      \\
2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3181       & 3200        & -0.04      & -0.01     & 0.01       \\
2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        \\
2M05030563+2122362 & 75.77354  & 21.37664  & 3155        & 3290       & 3503        & 0.06       & -0.15     & -0.36      \\
2M05222053+3031097 & 80.58561  & 30.51933  & 3389        & 3417       & 3426        & 0.28       & 0.2       & 0.14       \\
2M05312734-0340356 & 82.86417  & -3.67722  & 3801        & 3820       & 3854        & 0.49       & 0.49      & 0.37       \\
2M05413073+5329239 & 85.37804  & 53.48987  & 3765        & 3752       & 3755        & 0.19       & 0.15      & 0.13       \\
2M05420897+1229252 & 85.53833  & 12.48956  & 3250        & 3226       & 3235        & -0.22      & -0.34     & -0.31      \\
2M06000351+0242236 & 90.01458  & 2.70657   & 3214        & 3164       & 3165        & 0.07       & 0.05      & 0.03       \\
2M06011106+5935508 & 90.2961   & 59.59713  & 3340        & 3258       & 3255        & -0.09      & -0.05     & -0.03      \\
2M06112610+1032599 & 92.8588   & 10.54998  & 3636        & 3711       & 3698        & -0.39      & -0.42     & -0.39      \\
2M06544902+3316058 & 103.70411 & 33.26823  & 3448        & 3364       & 3359        & -0.02      & 0.04      & 0.06       \\
2M07171706-0501031 & 109.32108 & -5.01754  & 3193        & 3159       & 3168        & -0.09      & -0.12     & -0.09      \\
2M07272450+0513329 & 111.852   & 5.2263    & 3317        & 3285       & 3285        & -0.11      & -0.2      & -0.19      \\
2M07444018+0333089 & 116.16744 & 3.55252   & 3217        & 3176       & 3174        & 0.23       & 0.26      & 0.15       \\
2M08031949+5250387 & 120.83131 & 52.84402  & 3508        & 3622       & 3626        & -0.26      & -0.24     & -0.24      \\
2M08083284+5304377 & 122.13679 & 53.07712  & 3398        & 3280       & 3273        & 0.0        & 0.08      & 0.09       \\
2M08103429-1348514 & 122.64277 & -13.81427 & 3544        & 3517       & 3515        & 0.0        & 0.0       & 0.0        \\
2M08155393+3136392 & 123.97459 & 31.61077  & 3753        & 3810       & 3821        & 0.16       & 0.07      & 0.03       \\
2M08323599+4510175 & 128.14998 & 45.17153  & 3494        & 3505       & 3506        & -0.14      & -0.21     & -0.21      \\
2M08370799+1507475 & 129.28319 & 15.12944  & 3596        & 3553       & 3552        & -0.17      & -0.26     & -0.27      \\
2M08524084+2818589 & 133.17018 & 28.31634  & 3166        & 3213       & 3239        & 0.31       & 0.5       & 0.27       \\
2M08585633+0828259 & 134.73479 & 8.47384   & 3200        & 3142       & 3147        & -0.04      & -0.09     & -0.08      \\
2M09142298+5241125 & 138.595   & 52.68639  & 3920        & 3900       & 3879        & -0.01      & -0.02     & 0.01       \\
2M09174473+4612246 & 139.4363  & 46.20685  & 3478        & 3548       & 3556        & 0.04       & -0.02     & -0.03      \\
2M09214911+4330284 & 140.45447 & 43.50784  & 3170        & 3196       & 3206        & 0.16       & 0.23      & 0.19       \\
2M09285333-0722148 & 142.22221 & -7.37095  & 3508        & 3445       & 3440        & -0.27      & -0.23     & -0.22      \\
2M10112218+4927153 & 152.84254 & 49.45431  & 4131        & 4007       & 3946        & 0.24       & 0.22      & 0.23       \\
2M10121768-0344441 & 153.07364 & -3.74563  & 3623        & 3619       & 3620        & 0.13       & 0.16      & 0.16       \\
2M10193634+1952122 & 154.90117 & 19.87002  & 3370        & 3384       & 3389        & 0.15       & 0.24      & 0.21       \\
2M10285555+0050275 & 157.23155 & 0.8411    & 3548        & 3517       & 3515        & -0.18      & -0.16     & -0.15      \\
2M10331367+3409120 & 158.30687 & 34.15333  & 3500        & 3518       & 3518        & -0.15      & -0.19     & -0.2       \\
2M10350859+3349499 & 158.7858  & 33.83055  & 3816        & 3906       & 3907        & -0.04      & 0.01      & 0.03       \\
2M10453795+1833111 & 161.4081  & 18.55306  & 3673        & 3773       & 3780        & 0.14       & 0.13      & 0.13       \\
2M10505201+0648292 & 162.71704 & 6.80826   & 3238        & 3258       & 3265        & 0.16       & 0.18      & 0.12       \\
2M10520440+1359509 & 163.01762 & 13.99756  & 3298        & 3230       & 3228        & -0.13      & -0.11     & -0.07      \\
2M11000432+2249592 & 165.01781 & 22.833    & 3463        & 3426       & 3423        & -0.11      & -0.1      & -0.09      \\
2M11032023+3558117 & 165.83423 & 35.97054  & 3563        & 3682       & 3551        & -0.38      & -0.58     & -0.39      \\
2M11052903+4331357 & 166.36992 & 43.52665  & 3619        & 3547       & 3525        & -0.37      & -0.34     & -0.3       \\
2M11054316+1014093 & 166.42966 & 10.23605  & 3439        & 3395       & 3393        & -0.03      & -0.06     & -0.07      \\

>>>>

\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED BY DATA-DRIVEN MODELS}

@@

The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. We then use 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 68K and 0.07 dex respectively. Finally we apply our model to determine spectral subtypes, temperatures and metallicities for 10,000+ M dwarfs in the APOGEE-Gaia cross match.

@@

% \begin{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.

Define what parameters/classifications are important for M dwarfs. We want to estimate empirical parameters/classifications (SPT, etc.), physical parameters (such as Teff, logg, metallicity), which we will refer to as \emph{stellar labels}.

State why fundamental physical parameters are dificult to measure for M dwarfs. When temperatures and metallicities for a star are unavailable, it is often useful to classify a star in terms of spectral type. Define spectral type (this is a classification that is based on how a spectra looks, something to do with TiO band strengths, and it correlates with the temperature of a star).

% Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them:

Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.

The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.

A number of M dwarf studies have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity). \citealt{Gilhool:2018}; \citealt{Skinner:2018}.

Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.

@@

\subsection{Training Sample Selection}

The Cannon is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}). The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not. Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier or more accurate to infer from.

@@

\subsection{Test Sample Selection}

@@

APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample.

@@

The Cannon is a fully empirical model, which applies no use of line lists or radiative transfer models. Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label.

While The Cannon can in principle be trained on any physical or empirical labels.

Inferring the label of a star with such a model requires two steps: first the \emph{training step} in which a generative model is constructed at each pixel from the set of reference labels and fluxes, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum.

Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}. So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
\begin{equation}
	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise}
\end{equation}

Where the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, which the root mean sum of (1) the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and (2) the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$. Quadratic label vectors:
	\[\ell_{n} = [1, \, SPT, \, SPT^{2}] \]
	\[\ell_{n} = [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}] \]

The training step is the stage

@@

\subsection{Training Sample}

For each training sample

Two tests of consistency were used to assess the validity of our derived test labels (Figure \ref{fig:west_validation}). The self-test comparing training labels and test labels where precision and bias of the model are calculated as the standard deviation and mean of the difference in training and test labels respectively.

Some note on what lines vary significantly with parameter?

\subsubsection{Spectral Type Model}
We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources $<$M7, with only one M8 and one M9 source.

\subsubsection{Temperature/Metallicity Model}
For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the precision of the training uncertainties of 60K/0.08dex.

We note for this model XX out of 87 sources show strong rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), however the model

\subsection{Test Sample}

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=18cm]{figures/self_and_validation_test_spt_snr.png}
\end{center}
\caption{Consistency test for the West-trained spectral type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:west_validation}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=18cm]{figures/self_and_validation_test_teff_snr.png}
\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h_snr.png}
\end{center}
\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:mann_validation}
\end{figure}

%---------------------------
\begin{figure}[ht]
\plotone{figures/Spectral_Sequence_1.pdf}
\plotone{figures/Spectral_Sequence_2.pdf}
\plotone{figures/Spectral_Sequence_3.pdf}
\caption{ Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
\end{figure}

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities. \textit{Third panel:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}. \label{fig:demo_teff}
\end{figure}

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_feh1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures. \textit{Third panel:} Derivative of the Cannon model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
\end{figure}

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/derivative_jackknife_spt.png}
\end{center}
\caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$.} \label{fig:west_derivative}
\end{figure}

@@

%---------------------------
\begin{figure}[!htb]
\minipage{0.32\textwidth}
  \includegraphics[width=\linewidth]{figures/test_cmd_teff.png}
\endminipage\hfill
\minipage{0.32\textwidth}
  \includegraphics[width=\linewidth]{figures/test_cmd_feh.png}
\endminipage\hfill
\minipage{0.32\textwidth}%
  \includegraphics[width=\linewidth]{figures/test_cmd_chisq.png}
\endminipage
\end{figure}

@@

% \startlongtable
% \begin{deluxetable*}{ccc|ccc|ccc}
% \tablecaption{Cannon results for Temperature/Metallicity model \label{mann_results}}
% \tabletypesize{\scriptsize}
% \tablehead{
% \multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
% 	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}  
% } 
% \startdata
% 2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3532       & 3519        & -0.3       & -0.27     & -0.24      \\
% 2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3527       & 3554        & -0.3       & -0.28     & -0.27      \\
% 2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3185       & 3188        & 0.11       & 0.05      & 0.02       \\
% 2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3783       & 3778        & 0.04       & 0.11      & 0.13       \\
% 2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3094       & 3101        & -0.07      & -0.01     & 0.03       \\
% 2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3218       & 3217        & 0.1        & 0.03      & 0.0        \\
% 2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3057       & 3074        & -0.16      & -0.1      & -0.01      \\
% 2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3249       & 3253        & -0.12      & -0.29     & -0.29      \\
% 2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3468       & 3467        & -0.12      & -0.26     & -0.27      \\
% 2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3381       & 3369        & -0.35      & -0.25     & -0.18      \\
% 2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3181       & 3200        & -0.04      & -0.01     & 0.01       \\
% 2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        \\
% 2M05030563+2122362 & 75.77354  & 21.37664  & 3155        & 3290       & 3503        & 0.06       & -0.15     & -0.36      \\
% 2M05222053+3031097 & 80.58561  & 30.51933  & 3389        & 3417       & 3426        & 0.28       & 0.2       & 0.14       \\
% 2M05312734-0340356 & 82.86417  & -3.67722  & 3801        & 3820       & 3854        & 0.49       & 0.49      & 0.37       \\
% 2M05413073+5329239 & 85.37804  & 53.48987  & 3765        & 3752       & 3755        & 0.19       & 0.15      & 0.13       \\
% 2M05420897+1229252 & 85.53833  & 12.48956  & 3250        & 3226       & 3235        & -0.22      & -0.34     & -0.31      \\
% 2M06000351+0242236 & 90.01458  & 2.70657   & 3214        & 3164       & 3165        & 0.07       & 0.05      & 0.03       \\
% 2M06011106+5935508 & 90.2961   & 59.59713  & 3340        & 3258       & 3255        & -0.09      & -0.05     & -0.03      \\
% 2M06112610+1032599 & 92.8588   & 10.54998  & 3636        & 3711       & 3698        & -0.39      & -0.42     & -0.39      \\
% 2M06544902+3316058 & 103.70411 & 33.26823  & 3448        & 3364       & 3359        & -0.02      & 0.04      & 0.06       \\
% 2M07171706-0501031 & 109.32108 & -5.01754  & 3193        & 3159       & 3168        & -0.09      & -0.12     & -0.09      \\
% 2M07272450+0513329 & 111.852   & 5.2263    & 3317        & 3285       & 3285        & -0.11      & -0.2      & -0.19      \\
% 2M07444018+0333089 & 116.16744 & 3.55252   & 3217        & 3176       & 3174        & 0.23       & 0.26      & 0.15       \\
% 2M08031949+5250387 & 120.83131 & 52.84402  & 3508        & 3622       & 3626        & -0.26      & -0.24     & -0.24      \\
% 2M08083284+5304377 & 122.13679 & 53.07712  & 3398        & 3280       & 3273        & 0.0        & 0.08      & 0.09       \\
% 2M08103429-1348514 & 122.64277 & -13.81427 & 3544        & 3517       & 3515        & 0.0        & 0.0       & 0.0        \\
% 2M08155393+3136392 & 123.97459 & 31.61077  & 3753        & 3810       & 3821        & 0.16       & 0.07      & 0.03       \\
% 2M08323599+4510175 & 128.14998 & 45.17153  & 3494        & 3505       & 3506        & -0.14      & -0.21     & -0.21      \\
% 2M08370799+1507475 & 129.28319 & 15.12944  & 3596        & 3553       & 3552        & -0.17      & -0.26     & -0.27      \\
% 2M08524084+2818589 & 133.17018 & 28.31634  & 3166        & 3213       & 3239        & 0.31       & 0.5       & 0.27       \\
% 2M08585633+0828259 & 134.73479 & 8.47384   & 3200        & 3142       & 3147        & -0.04      & -0.09     & -0.08      \\
% 2M09142298+5241125 & 138.595   & 52.68639  & 3920        & 3900       & 3879        & -0.01      & -0.02     & 0.01       \\
% 2M09174473+4612246 & 139.4363  & 46.20685  & 3478        & 3548       & 3556        & 0.04       & -0.02     & -0.03      \\
% 2M09214911+4330284 & 140.45447 & 43.50784  & 3170        & 3196       & 3206        & 0.16       & 0.23      & 0.19       \\
% 2M09285333-0722148 & 142.22221 & -7.37095  & 3508        & 3445       & 3440        & -0.27      & -0.23     & -0.22      \\
% 2M10112218+4927153 & 152.84254 & 49.45431  & 4131        & 4007       & 3946        & 0.24       & 0.22      & 0.23       \\
% 2M10121768-0344441 & 153.07364 & -3.74563  & 3623        & 3619       & 3620        & 0.13       & 0.16      & 0.16       \\
% 2M10193634+1952122 & 154.90117 & 19.87002  & 3370        & 3384       & 3389        & 0.15       & 0.24      & 0.21       \\
% 2M10285555+0050275 & 157.23155 & 0.8411    & 3548        & 3517       & 3515        & -0.18      & -0.16     & -0.15      \\
% 2M10331367+3409120 & 158.30687 & 34.15333  & 3500        & 3518       & 3518        & -0.15      & -0.19     & -0.2       \\
% 2M10350859+3349499 & 158.7858  & 33.83055  & 3816        & 3906       & 3907        & -0.04      & 0.01      & 0.03       \\
% 2M10453795+1833111 & 161.4081  & 18.55306  & 3673        & 3773       & 3780        & 0.14       & 0.13      & 0.13       \\
% 2M10505201+0648292 & 162.71704 & 6.80826   & 3238        & 3258       & 3265        & 0.16       & 0.18      & 0.12       \\
% 2M10520440+1359509 & 163.01762 & 13.99756  & 3298        & 3230       & 3228        & -0.13      & -0.11     & -0.07      \\
% 2M11000432+2249592 & 165.01781 & 22.833    & 3463        & 3426       & 3423        & -0.11      & -0.1      & -0.09      \\
% 2M11032023+3558117 & 165.83423 & 35.97054  & 3563        & 3682       & 3551        & -0.38      & -0.58     & -0.39      \\
% 2M11052903+4331357 & 166.36992 & 43.52665  & 3619        & 3547       & 3525        & -0.37      & -0.34     & -0.3       \\
% 2M11054316+1014093 & 166.42966 & 10.23605  & 3439        & 3395       & 3393        & -0.03      & -0.06     & -0.07      \\
% 2M11091225-0436249 & 167.30106 & -4.60695  & 3661        & 3743       & 3748        & -0.03      & -0.09     & -0.1       \\
% 2M11263270+2145262 & 171.63622 & 21.75719  & 3610        & 3642       & 3642        & -0.18      & -0.16     & -0.15      \\
